---
title: "BIOS 643 - Health Data Analysis: Thry to Prac - Analysis & Rptg"
author: "William Opoku-Nkoom"
subtitle: Assessment Week 2
output:
  pdf_document: default
  html_document:
    df_print: paged
header-includes:
- \usepackage{titling}
- \pretitle{\begin{center} \includegraphics{McG_DEPT EPIDEM BIOSTAT OCCUPAT HEALTH_rgb_red_hor_en.jpg}\LARGE\\}
- \posttitle{\end{center}}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r echo = FALSE}
# load relevant packages
pacman::p_load(rio,
               here,
               ggplot2,
               dplyr,
               tidyverse,
               skimr,
               lubridate,
               janitor
               )

```


```{r include = FALSE}
# load cohort dataset 
cohort <- import(here("BIOS643", "Week_2", "Assessment_2", "cohort-data.csv"))

# data overview
head(cohort)
tail(cohort)
skim(cohort) 
str(cohort)
names(cohort)
```

# Data Cleaning Plan

**Cohort Dataset**

*Issues found from data overview and how corrected*

 * The variables entry date and lost to follow-up date had some values that could not be recognized as dates.
   * The parse_date_time() function of the lubridate package was used to identify and validate all date formats.
   * The original entry date and lost to follow-up variables were then dropped after creating new variables using the parse_date_time() function of the lubridate package.
* Column names were inconsistent with a mix of lower and upper cases with no structured format
  * The dplyr package was used to rename specific columns
  * The janitor package was used to standardize variable names
  * Columns were finally logically arranged with the select() function
* Duplicated row check
  * Three (3) duplicated rows found; specifically id: 38, 1782 and 1744
  * These were then removed to have distinct rows
* The skim() function from the skimr package showed no missing values
* The variables id, age, sex, bmi are numeric from the skim() function of the skimr package
  * sex was therefore recoded as factor
* Smoking status was not accurately labelled (mix of upper and lower cases)
  * Mutated to have unique labels and recoded as a factor
* Body mass index (bmi) was categorised into underweight, normal, overweight and obese and then orderd as a factor variable
* Socio-economic status (ses_status) was also ordered as as factor variable
* Application of the minimum age at entry excluded 8 participants from the cohort.
* The cleaned cohort dataset was then exported for sharing and future analysis.

```{r echo = FALSE, warning = FALSE}
# filter rows with entry dates and loss to follow-up dates
cohort$entry_date_parsed <- parse_date_time(
  cohort$entrydate, 
  orders = c("ymd", "dmy", "mdy", "Ymd", "dmY", "mdY")
  )

cohort$date_loss_parsed <- parse_date_time(
  cohort$DATE_losstofollowup, 
  orders = c("ymd", "dmy", "mdy", "Ymd", "dmY", "mdY")
  )

cohort <- cohort %>%
  filter(!is.na(entry_date_parsed)) # drops NA dates

# drop original entry date and lost to follow-up variables
cohort <- cohort %>%
  select(-entrydate, -DATE_losstofollowup)

# clean column names
cohort <- cohort %>%
  rename(
    entry_date = entry_date_parsed,
    sex = `SEX (1=MALE)`,
    ses_status = socioeconomic_status,
    smoking_status = SMOKE,
    lost_date = date_loss_parsed
  )

cohort <- clean_names(cohort) # standardize column names

# recode sex as a factor
cohort$sex <- factor(cohort$sex, levels = c(0, 1),
                     labels = c("Female", "Male"))

# standardize smoking_status levels then recode as factor
cohort$smoking_status <- tolower(cohort$smoking_status)
cohort$smoking_status <- factor(
  tools::toTitleCase(cohort$smoking_status),
  levels = c("Never", "Former", "Current")
)
table(cohort$smoking_status) # check

# categorise bmi
cohort <- cohort %>%
  mutate(bmi_cat = case_when(
    bmi < 18.5 ~ "Underweight",
    bmi < 25 ~ "Normal",
    bmi < 30 ~ "Overweight",
    TRUE ~ "Obese"
  )
  )
table(cohort$bmi_cat) # check

cohort$bmi_cat <- factor(
  cohort$bmi_cat, levels = c(
    "Underweight", "Normal", "Overweight", "Obese"
  )
) # ordered factor for bmi_cat

cohort$ses_status <- factor(
  cohort$ses_status, levels = c(
    "Low", "Medium", "High"
  )
) # ordered factor for ses_status

table(cohort$ses_status) # check

# arrange columns
cohort <- cohort %>%
  select(id, entry_date, age, sex, bmi, bmi_cat, ses_status, smoking_status, lost_date)

# check for duplicate rows
any(duplicated(cohort)) # True
sum(duplicated(cohort)) # 3 duplicated rows present
cohort[duplicated(cohort$id), ] # ids: 38, 1782, 1744 duplicated
cohort <- cohort[!duplicated(cohort$id), ] # remove duplicated rows

# application of exclusion criterion of 40 years minimum age in cohort
summary(cohort$age) # minimum age = 33 years
cohort_40 <- cohort %>%
  filter(age >= 40) # 8 participants excluded for their age of entry

# ensure all dates are date objects
cohort_40$entry_date <- as.Date(cohort_40$entry_date)
cohort_40$lost_date <- as.Date(cohort_40$lost_date)


# export cleaned dataset
export(cohort_40, here("BIOS643", "Week_2", "Assessment_2", "cohort_clean.csv"))


```

\newpage

**icd dataset**

*Issues found from data overview and how corrected*

* The variable "description" was renamed to icd_description
* Duplicated row check showed no duplicated rows
* skim() function from the skimr package showed no missing values
* Multiple diagnostic dates (icd_date) for the same person (id) for the same disease(icd_code):
  * slice_min() function of the dplyr package was used to only retain the first diagnosis. 
* COPD outcome indicator was created using the mutate() function and recoded as a factor, where 0 means copd absent (No) and 1 means copd present (Yes)
* A copd comorbidy indicator was also defined using the summary() function.
  * The comorbidity variable was recoded as a factor where 0 means absence of comorbidity (No) and 1 means the presence of comorbidity (Yes)
  * This mini dataset was then merged to the original to have a composite icd dataset.
* The cleaned icd dataset was then exported for sharing and future analysis

```{r include = FALSE}
# load icd dataset
icd <- import(here("BIOS643", "Week_2", "Assessment_2", "icd-data.csv"))

# data overview
head(icd)
tail(icd)
skim(icd) 
str(icd)
names(icd)

table(icd$description) # verify the various icd descriptions

```



```{r echo = FALSE}

icd <- icd %>%
  rename(
    icd_description = description # rename description as icd_description
    )
# check for duplicates
any(duplicated(icd)) # False

# keep only first diagnosis per person per disease
icd <- icd %>%
  group_by(id, icd_code) %>%
  slice_min(order_by = icd_date, with_ties = FALSE) %>%
  ungroup()

# create copd indicator
copd_pattern <- "^J44.0|^J44.1|^J44.9"
icd <- icd %>%
  mutate(
    copd_outcome = if_else(
      str_detect(icd_code, copd_pattern),
      1, 0
    )
  )

# recode copd_outcome as a factor
icd$copd_outcome <- factor(icd$copd_outcome, levels = c(0, 1),
                     labels = c("No", "Yes"))
table(icd$copd_outcome) # check

# flag co-morbidity
comorbidity <- icd %>%
  mutate(
    copd_flag = str_detect(icd_code, copd_pattern)
  ) %>%
  group_by(id) %>%
  summarise(
    copd_comorb = as.integer(
      any(copd_flag) & n() > 1
    )
  )
table(comorbidity$copd_comorb) # check

# recode copd_comorbidity as a factor
comorbidity$copd_comorb <- factor(comorbidity$copd_comorb, 
                                   levels = c(0, 1),
                     labels = c("No", "Yes"))
table(comorbidity$copd_comorb) # check

# merge comorbidity to icd dataset
icd_comob <- left_join(icd, comorbidity, by = "id")

# export cleaned dataset
export(icd_comob, here("BIOS643", "Week_2", "Assessment_2", "icd_clean.csv"))


```

\newpage


**residential dataset**

*Issues found from data overview and how corrected*

* skim() function from the skimr package showed no missing values
* Duplicated row check showed no duplicated rows
* All dates were ensured they were date objects from the as.Date() function of the lubridate package
* First residential address dates were grouped by particpant id and then summerised to calculate the number of years between first residential address date and cohort entry date (lag_years)
* The mini dataset was then merged to the original residential dataset, selecting only id and lag_years variables from the mini dataset
* Application of a minimum of 10 year lag_date (i.e period between participant's first residential address and cohort entry date) did not exclude any participant, as the minimum number of years found was 10.10 years from summary statistics.
* Variables were finally arranged in a logical format
* The cleaned residential dataset was was exported for sharing and future analysis.

```{r include = FALSE}
# load residential dataset
residential <- import(here("BIOS643", "Week_2", "Assessment_2", "res-data.csv"))

head(residential)
tail(residential)
skim(residential) 
str(residential)
names(residential)

```


```{r echo = FALSE, message = FALSE, warning = FALSE}

# check for duplicates
any(duplicated(residential)) # False

# calculate lag years between address start date and entry date

# ensure all dates are date objects
residential$address_start_date <- as.Date(residential$address_start_date)
residential$entry_date <- as.Date(residential$entry_date)
residential$address_end_date <- as.Date(residential$address_end_date)

# get first address date per participant
first_address <- residential %>%
  group_by(id) %>%
  summarise(
    first_address = min(address_start_date, na.rm = TRUE),
    entry_date = first(entry_date) 
  )

# calculate years between first address and cohort entry date
first_address <- first_address %>%
  mutate(
    lag_years = as.numeric(
      difftime(entry_date, first_address, units = "days")
    )/365.25
  )
# merge back to original dataset
residential <- residential %>%
  left_join(
    first_address %>%
      select(id, lag_years),
    by = "id"
  )
names(residential)

# rearrange columns
residential <- residential %>%
  select(id, address_start_date, entry_date, lag_years, address_end_date, 
         county, state)
summary(residential$lag_years) # min yrs btn first res add & entry date = 10.10 

# export cleaned dataset
export(residential, here("BIOS643", "Week_2", "Assessment_2", "res_clean.csv"))
```

\newpage


**Combined dataset**

* The as.Date() function of the lubridate package was used to ensure all dates are date objects
* Duplicated row check showed no duplicated rows
* The cleaned cohort dataset was first merged with the cleaned residential dataset using id and entry dates as key
* The resulting dataset was then merged with the cleaned icd dataset using id as key
* The variables were then rearranged using the select() function in a logical order
* Some variable names were cleaned and standardized
* skim() function from the skimr package showed that there are a total of 90 missing values from the following variables: addr_start_date, addr_end_date and lag_years.
* The final cleaned and merged dataset was exported for sharing and further analysis.

```{r include = FALSE}

# datasets overview 
dim(cohort)
dim(icd)
dim(residential)
names(cohort)
names(icd)
names(residential)

```


```{r echo = FALSE, warning = FALSE}
# ensure all dates are date objects
cohort$entry_date <- as.Date(cohort$entry_date)
cohort$lost_date <- as.Date(cohort$lost_date)
residential$address_start_date <- as.Date(residential$address_start_date)
residential$entry_date <- as.Date(residential$entry_date)
residential$address_end_date <- as.Date(residential$address_end_date)
icd_comob$ehr_entry_date <- as.Date(icd$ehr_entry_date)
icd_comob$icd_date <- as.Date(icd$icd_date)

# merge cohort and residential datasets
cohort_res <- left_join(cohort, residential, 
                            by = c("id", "entry_date"))
# check for duplicates
any(duplicated(cohort_res)) # False

# merge cohort_res and icd datasets
cohort_res_icd <- left_join(cohort_res, icd_comob, 
                            by = c("id")
                            )
# check duplicates
any(duplicated(cohort_res_icd)) # False

# arrange varriables
cohort_res_icd <- cohort_res_icd %>%
  select(id, entry_date, lag_years, age, sex, bmi, bmi_cat, ses_status,
         smoking_status,lost_date,county, state, address_start_date,
         address_end_date, ehr_entry_date,icd_code, 
         icd_description, icd_date, copd_outcome, copd_comorb)

# rename some variables
cohort_res_icd <- cohort_res_icd %>%
  rename(
    addr_start_date = address_start_date,
    addr_end_date = address_end_date
    )
names(cohort_res_icd) # check

```


```{r include = FALSE}
skim(cohort_res_icd)

# export final cleaned and merged dataset
export(cohort_res_icd, here("BIOS643", "Week_2", "Assessment_2", "cohort_res_icd.csv"))

```


\newpage


**Quality control checks**

1) Analysability
  * First role has only variables
  * There is participant id which serves as a unique identifier of participants
2) Interpretability
  * All variable names are unique and clearly readable
  * The longest variable has 16 characters which by far less than the maximum required number of character (32)
3) Completeness
  * No missingness
  * sample size = 1764, which is consistent from the output from the dim() function
  * Distinct rows (i.e. no duplicated values)
4) Validity
  * Date object have been ascertained
  * Numeric variables are numeric as expected
  * Categorical variables and factors are as expected
5) Accuracy
  * Overview of the data for now shows accuracy
6) Consistency
  * Value labeles like dates are uniformly applied, as well as categorical variables.'
7) De-identification
  * As it stands now, it is not possible to identify any specific individuals from the dataset.

  
\newpage

**Data dictionary**

Variable name  |  Data type  |    Values          | Notes                    |
---------------|-------------|--------------------|--------------------------|
id             |Numeric      | 1 to inf           |Unique identifier         |
---------------|-------------|--------------------|--------------------------|
entry_date     |Date         |YY-MM-DD            |Date of cohort entry      |
---------------|-------------|--------------------|--------------------------|
age            |Numeric      |18 and above        |Age of the respondant     |
---------------|-------------|--------------------|--------------------------|
sex            |Factor       |0=Female; 1=Male    |Biological sex            |
---------------|-------------|--------------------|--------------------------|
bmi            |Numeric      |17 and above        |Body mass index           |
---------------|-------------|--------------------|--------------------------|
bmi_cat        |Factor       |Underweight         |bmi < 18.5                |
---------------|-------------|--------------------|--------------------------|
.              | .           |Normal              |bmi 18.5 -24.9            |
---------------|-------------|--------------------|--------------------------|
.              | .           |Overweight          |bmi 25-29.9               |
---------------|-------------|--------------------|--------------------------|
.              | .           |Obese               |bmi > 30.0                |
---------------|-------------|--------------------|--------------------------|
ses_status     |Factor       |High/Medium/Low     |Socio-economic status     |
---------------|-------------|--------------------|--------------------------|
smoking_status |Factor       |Never/Former/Current|Smoking status            |
---------------|-------------|--------------------|--------------------------|
lost_date      |Date         |YY-MM-DD            |Date lost to follow-up    |
---------------|-------------|--------------------|--------------------------|
county         |Character    |US counties         |Counties in the US        |
---------------|-------------|--------------------|--------------------------|
state          |Character    |US states           |States in the US          |
---------------|-------------|--------------------|--------------------------|
addr_start_date|Date         |YY-MM-DD            |Start date of residence   |
---------------|-------------|--------------------|--------------------------|
addr_end_date  |Date         |YY-MM-DD            |End date of residence     |
---------------|-------------|--------------------|--------------------------|
ehr_entry_date |Date         |YY-MM-DD            |EHR entry date record     |
---------------|-------------|--------------------|--------------------------|
icd_code       |Character    |Alphanumeric        |Disease diagnostic codes  |
---------------|-------------|--------------------|--------------------------|
icd_description|Character    |-                   |Disease condition         |
---------------|-------------|--------------------|--------------------------|
icd_date       |Date         |YY-MM-DD            |Disease diagnosis date    |
---------------|-------------|--------------------|--------------------------|
copd_outcome   |Factor       |0=No; 1=Yes         |COPD outcome diagnosis    |
---------------|-------------|--------------------|--------------------------|
copd_comorb    |Factor       |0=No; 1=Yes         |COPD comorbidity diagnosis|
---------------|-------------|--------------------|--------------------------|
lag_years      |Numeric      |integer             |First residence-entry date|

 

              


